---
title: "Find genes that are friendly to a latent pattern (or a cluster) in a matrix"
output: 
  html_vignette:
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
 %\VignetteIndexEntry{Find genes that are friendly to a latent pattern (or a cluster) in a matrix}
---

[Naturam expellas furca, tamen usque recurret]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(rmarkdown)`
`r library(friends.test)`
`r options(width=80,keep.source=TRUE)`
`r options(timeout=300)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`
`r library('fgsea')`
`r library('msigdbr')`
<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">

# Introduction

## Genes that are friends to CoGAPS patterns (or CoGAPS patterns that are friends to genes?)

CoGAPS decomoses an expression dataset into pattetns of expression, ideally, eah of them represent an underlying biological process. We use the [PDAC dataset](https://zenodo.org/records/7709664) from the [Johnson _et al_, 2023](https://www.nature.com/articles/s41596-023-00892-x) as an example. 

    ```{r 'setup'}
    data(friends.test.cogaps.example)
    ```

Feature loadings in CoGAPS are normalized expression values across patterns. Each column is a pattern, each row is a gene. 

    ```{r 'examine input'}
    featureLoadings <- friends.test.cogaps.example$loadings
    head(featureLoadings)
    ```

Let's look for marker genes for the patterns using `friends.test`. Genes are rows, patterns are columns. Let's looki for genes that have only one friend (exclusive markers), so we use `max.friends.n = 1`.

    ```{r 'patterns are genes friends'}
    best_friends_of_genes <- friends.test(featureLoadings, max.friends.n = 1)
    names(best_friends_of_genes)
    ```
Each of the genes is a marker of a pattern. The output format allows a marker to have a lot of friends. It is list by marker, and each element is list by friend, and the leave is a 3-integer vector:

    ```{r show list element}
    best_friends_of_genes[[1]][[1]]
    ```
Thte three integers are the marker index, friend index, and the rank of the friend for the marker. Now, the third is always 1. The format is flexible enough.

For our example, let's pack it into a data frame. It works ok if the number of elements do not reach hundred of thousands.

    ```{r 'display all genes per best friend'}
    best_friends_df <- best_friends_of_genes |>
        purrr::flatten() |>
        purrr::reduce(.f = rbind) |>
        data.frame() |>
        tibble::remove_rownames() |>
        dplyr::mutate(
            gene = rownames(featureLoadings)[marker],
            pattern = colnames(featureLoadings)[friend]
        )
    head(best_friends_df)
    ```

Now, let's prepare a list of markers by pattern.

    ```{r 'make exclusive marker list'}
    exclusive_markers_list <- best_friends_df |>
        split(best_friends_df$pattern) |>
        purrr::map(~ .x |>
                dplyr::pull(gene) |>
                unique() |>
                sort()
        )
    ```

Lets examine hallmarks associated with list of the pattern markers. We can use `fgsea` for that. We will use the loadings as a ranking statistic, and we will look up the hallmark gene sets from MSigDB.

    ```{r 'signatures in one-hot lists of exclusive markers'}
    best_friend_patterns <- best_friends_df$pattern |> unique() |> sort()
    msigdbr_df <- msigdbr::msigdbr(species = "human", collection = "H")
    pathways <- split(
                msigdbr_df$gene_symbol,
                msigdbr_df$gs_name)
    hallmarks_by_exclusive_markers <- lapply(best_friend_patterns, function(pat) {
        markers <- best_friends_df |>
            dplyr::filter(pattern == pat) |>
            dplyr::pull(gene)

        fgseaRes <- fgsea::fora(
            pathways = pathways,
            markers,
            rownames(featureLoadings)
        )
        cat(markers)
        #fgseaRes[padj < 0.05, c('pathway', 'padj')][order(padj)]
        fgseaRes
    })
    hallmarks_by_exclusive_markers
    ```

Now, let's create lists of markers per patter.

    ```{r 'list of marker per best friend', eval = FALSE}

    #create a list of friends (gene names) per pattern with purrr
    friends_list <- friends_tibble |>
        split(friends_tibble$friend) |>
        purrr::map(~rownames(featureLoadings)[.$marker]) |>
        (\(x) setNames(x, colnames(featureLoadings)[as.integer(names(x))]))()
    ```

Lets examine hallmarks associated with patterns. We can use `fgsea` for that. We will use the loadings as a ranking statistic, and we will look up the hallmark gene sets from MSigDB.

    ```{r 'look up signatures in CoGAPS patterns', eval=FALSE}
library(fgsea)
library(msigdbr)
msigdbr_df <- msigdbr(species = "human", category = "H")

# 1. run for CoGAPS patterns
hallmarks <- lapply(names(friends_list), function(pat) {
    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                     stats = featureLoadings[, pat])
    fgseaRes[padj < 0.05, c('pathway', 'padj')][order(padj)]

}) |> setNames(names(friends_list))

hallmarks
    ```

    ```{r 'signatures in one-hot ordered lists',eval=FALSE}

# 2. run with one-hot friends vs not friends as ranking

hallmarks_onehot <- lapply(names(friends_list), function(pat) {

    onehot <- names(featureLoadings[, pat]) %in% friends_list[[pat]] |>
        as.numeric() |>
        setNames(names(featureLoadings[, pat]))

    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                      stats = onehot)
    fgseaRes[padj < 0.05, c('pathway', 'padj')][order(padj)]

}) |> setNames(names(friends_list))

hallmarks_onehot
    ```

    ```{r 'signatures in pattern loadings ordered by friends lists', eval=FALSE}

# 3. run with loadings but friends are ranked higher than non-friends

ratings <- matrix(0, nrow(featureLoadings), ncol(featureLoadings))

for (i in seq_len(nrow(friends_tibble))){
    ratings[unlist(friends_tibble[i, 1]),
            unlist(friends_tibble[i, 2])] <- 1 / unlist(friends_tibble[i, 3])
}

colnames(ratings) <- colnames(featureLoadings)
rownames(ratings) <- rownames(featureLoadings)

hallmarks_rated <- lapply(names(friends_list), function(pat) {
    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                     stats = ratings[, pat])
    fgseaRes[padj < 0.05, c("pathway", "padj")][order(padj)]

}) |> setNames(names(friends_list))

hallmarks_rated
    ```

    ```{r 'signatures in ordered friends lists', eval=FALSE}

# 4. hallmarks directly on friends results

hallmarks_friends <- lapply(names(friends_list), function(pat) {
    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                      stats = featureLoadings[friends_list[[pat]], pat])
    fgseaRes[padj < 0.05, c("pathway", "padj")][order(padj)]

}) |> setNames(names(friends_list))

hallmarks_friends

    ```


assess marker specificity (are friends of pattern 1 also friends of pattern 2?)

    ```{r 'specificity measure'}
    #for each pattern, find avarage rank of friends in that pattern vs average rank in other patterns
    specificity <- lapply(names(friends_list), function(pat) {
        friends <- friends_list[[pat]]
        friends_ranks_in_pattern <- rank(featureLoadings[, pat])[friends]
        avg_rank_friends <- mean(friends_ranks_in_pattern)
        others <- setdiff(names(featureLoadings[, pat]), friends)
        others_ranks_in_pattern <- rank(featureLoadings[, pat])[others]
        avg_rank_others <- mean(others_ranks_in_pattern)
        rank_friends_in_other_patterns <- sapply(setdiff(colnames(featureLoadings), pat), function(other_pat) {
            rank(featureLoadings[, other_pat])[friends]
        })
        avg_rank_friends_in_other_patterns <- apply(rank_friends_in_other_patterns, 1, mean) |> mean()
        list(pattern = pat,
            avg_rank_friends = avg_rank_friends,
            avg_rank_others = avg_rank_others,
            avg_rank_friends_in_other_patterns = avg_rank_friends_in_other_patterns,
            count_friends = length(friends)
        )
    })
    specificity
    ```

HERE! - Next steps: 
    - run on horizontal matrix

Compare all pattern friends to CoGAPS pattern markers. All CoGAPS pattern markers are friends, but not all friends are CoGAPS pattern markers.

    ```{r 'compare with CoGAPS pattern markers', eval=FALSE}
    pm_all <- friends.test.cogaps.example$markers_all[["PatternMarkers"]]
    comp <- lapply(names(pm_all), function(pat) {
        list(pattern = pat,
            pm_in_friends = pm_all[[pat]] %in% names(friends_genes[[pat]]) |> mean(),
            friends_in_pm = names(friends_genes[[pat]]) %in% pm_all[[pat]] |> mean())
    })
    do.call(rbind, comp)
    ```


Compare with CoGAPS pattern markers again.
    ```{r 'compare with CoGAPS pattern markers again', eval=FALSE}
    comp <- lapply(names(pm_all), function(pat){
        list(pattern = pat,
            pm_in_friends = pm_all[[pat]] %in% by_patterns[[pat]] |> mean(),
            friends_in_pm = by_patterns[[pat]] %in% pm_all[[pat]] |> mean()
        )
    })
    do.call(rbind, comp)
    ```

</div>
