---
title: "Find genes that are friendly to a latent pattern (or a cluster) in a matrix"
output: 
  html_vignette:
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
 %\VignetteIndexEntry{Find genes that are friendly to a latent pattern (or a cluster) in a matrix}
---

[Naturam expellas furca, tamen usque recurret]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(rmarkdown)`
`r library(friends.test)`
`r options(width=80,keep.source=TRUE)`
`r options(timeout=300)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`
`r devtools::load_all()` #remove me
<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">

# Introduction

## Genes that are friends to CoGAPS patterns (or CoGAPS patterns that are friends to genes?)

```{r 'setup'}
data(friends_test_cogaps_example)
```

Feature loadings in CoGAPS are normalized expression values across patterns. Each column is a pattern, each row is a gene.
```{r 'examine input'}
head(featureLoadings)
```

Let's find representative genes for all patterns using `friends.test`. We are interested in gene-friends for all patterns, so patterns need to be in rows, and genes in columns.
```{r 'friends are genes'}
friends_genes <- friends.test(t(featureLoadings))
```

Wow, there are so many friends for each pattern! (TODO Write here what we found)
```{r 'friends summary'}
lapply(friends_genes, length)
```

Now let's inspect loadings values for Pattern4 friends vs non-friends. Higher loading for friend genes is observed, but there is no clear cut-off. 
In the highest loading value decile, about 50% of genes are friends, but also in the lowest decile, there are about 2% of friend genes.
This is because ... TODO
```{r 'inspect pattern 4 friends'}
loadings <- featureLoadings[, "Pattern_4"]
df_ecdf <- ecdf(loadings)
df <- data.frame(value = loadings,
                 friend = names(loadings) %in% names(friends_genes$Pattern_4),
                 loading_ecdf = df_ecdf(loadings))

aggregate(friend ~ round(loading_ecdf, 1), data = df, 
          FUN = function(x) c(share = mean(x), n = sum(x)))
```

Compare all pattern friends to CoGAPS pattern markers. All CoGAPS pattern markers are friends, but not all friends are CoGAPS pattern markers.
```{r 'compare with CoGAPS pattern markers'}
comp <- lapply(names(pm_all), function(pat) {
    list(pattern = pat,
        pm_in_friends = pm_all[[pat]] %in% names(friends_genes[[pat]]) |> mean(),
        friends_in_pm = names(friends_genes[[pat]]) %in% pm_all[[pat]] |> mean())
})
do.call(rbind, comp)
```

TODO Validate! Lets find very specific genes for each pattern (genes that have only one friend). We can call them "pattern-specific genes".
```{r 'fiends are patterns'}
friends_patterns <- friends.test(featureLoadings, max.friends.n = 1)

qq <- lapply(friends_patterns, function(x) names(x)) |> as.data.frame() |> t()

by_patterns <- lapply(unique(qq), function(pat) rownames(qq)[qq == pat])
names(by_patterns) <- unique(qq)

by_patterns
```

Compare with CoGAPS pattern markers again!
```{r 'compare with CoGAPS pattern markers again'}
comp <- lapply(names(pm_all), function(pat){
    list(pattern = pat,
         pm_in_friends = pm_all[[pat]] %in% by_patterns[[pat]] |> mean(),
         friends_in_pm = by_patterns[[pat]] %in% pm_all[[pat]] |> mean())
})
do.call(rbind, comp)
```
</div>