---
title: "Find genes that are friendly to a latent pattern (or a cluster) in a matrix"
output: 
  html_vignette:
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
 %\VignetteIndexEntry{Find genes that are friendly to a latent pattern (or a cluster) in a matrix}
---

[Naturam expellas furca, tamen usque recurret]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(rmarkdown)`
`r library(friends.test)`
`r options(width=80,keep.source=TRUE)`
`r options(timeout=300)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`
`r library('fgsea')
<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">

# Introduction

## Genes that are friends to CoGAPS patterns (or CoGAPS patterns that are friends to genes?)

    ```{r 'setup'}
    data(friends.test.cogaps.example)
    ```

Feature loadings in CoGAPS are normalized expression values across patterns. Each column is a pattern, each row is a gene.

    ```{r 'examine input'}
    featureLoadings <- friends.test.cogaps.example$loadings
    head(featureLoadings)
    ```

Let's find representative genes for all patterns using `friends.test`. We are interested in patterns - friends of genes, so genes need to be in rows, and patterns in columns.

    ```{r 'patterns are genes friends'}
    friends_of_genes <- friends.test(featureLoadings)
    ```


    ```{r 'display all genes per friend'}
    friends_tibble <- friends_of_genes |>
        purrr::flatten() |>
        purrr::reduce(.f = rbind) |>
        tibble::as_tibble()

    #create a list of friends (gene names) per pattern with purrr
    friends_list <- friends_tibble |>
        split(friends_tibble$friend) |>
        purrr::map(~rownames(featureLoadings)[.$marker]) |>
        setNames(featureLoadings |> colnames())
    ```

Lets examine hallmarks associated with patterns. We can use `fgsea` for that. We will use the loadings as a ranking statistic, and we will look up the hallmark gene sets from MSigDB.

    ```{r 'look up signatures in CoGAPS patterns'}
library(fgsea)
library(msigdbr)
msigdbr_df <- msigdbr(species = "human", category = "H")

# 1. run for CoGAPS patterns
hallmarks <- lapply(names(friends_list), function(pat) {
    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                     stats = featureLoadings[, pat])
    fgseaRes[padj < 0.05, c('pathway', 'padj')][order(padj)]

}) |> setNames(names(friends_list))

hallmarks
    ```

    ```{r 'signatures in one-hot ordered lists'}

# 2. run with one-hot friends vs not friends as ranking

hallmarks_onehot <- lapply(names(friends_list), function(pat) {

    onehot <- names(featureLoadings[, pat]) %in% friends_list[[pat]] |>
        as.numeric() |>
        setNames(names(featureLoadings[, pat]))

    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                      stats = onehot)
    fgseaRes[padj < 0.05, c('pathway', 'padj')][order(padj)]

}) |> setNames(names(friends_list))

hallmarks_onehot
    ```

    ```{r 'signatures in pattern loadings ordered by friends lists'}

# 3. run with loadings but friends are ranked higher than non-friends

ratings <- matrix(0, nrow(featureLoadings), ncol(featureLoadings))

for (i in seq_len(nrow(friends_tibble))){
    ratings[unlist(friends_tibble[i, 1]),
            unlist(friends_tibble[i, 2])] <- 1 / unlist(friends_tibble[i, 3])
}

colnames(ratings) <- colnames(featureLoadings)
rownames(ratings) <- rownames(featureLoadings)

hallmarks_rated <- lapply(names(friends_list), function(pat) {
    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                     stats = ratings[, pat])
    fgseaRes[padj < 0.05, c("pathway", "padj")][order(padj)]

}) |> setNames(names(friends_list))

hallmarks_rated
    ```

    ```{r 'signatures in ordered friends lists'}

# 4. hallmarks directly on friends results

hallmarks_friends <- lapply(names(friends_list), function(pat) {
    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                       msigdbr_df$gs_name),
                      stats = featureLoadings[friends_list[[pat]], pat])
    fgseaRes[padj < 0.05, c("pathway", "padj")][order(padj)]

}) |> setNames(names(friends_list))

hallmarks_friends

    ```


assess marker specificity (are friends of pattern 1 also friends of pattern 2?)

    ```{r 'specificity measure'}
    #for each pattern, find avarage rank of friends in that pattern vs average rank in other patterns
    specificity <- lapply(names(friends_list), function(pat) {
        friends <- friends_list[[pat]]
        friends_ranks_in_pattern <- rank(featureLoadings[, pat])[friends]
        avg_rank_friends <- median(friends_ranks_in_pattern)
        others <- setdiff(names(featureLoadings[, pat]), friends)
        others_ranks_in_pattern <- rank(featureLoadings[, pat])[others]
        avg_rank_others <- median(others_ranks_in_pattern)
        list(pattern = pat,
             avg_rank_friends = avg_rank_friends,
             avg_rank_others = avg_rank_others,
             specificity_score = avg_rank_friends/avg_rank_others)
    })
    specificity
    ```

HERE! - Next steps: 
    - run on horizontal matrix

Compare all pattern friends to CoGAPS pattern markers. All CoGAPS pattern markers are friends, but not all friends are CoGAPS pattern markers.

    ```{r 'compare with CoGAPS pattern markers'}
    pm_all <- friends.test.cogaps.example$markers_all[["PatternMarkers"]]
    comp <- lapply(names(pm_all), function(pat) {
        list(pattern = pat,
            pm_in_friends = pm_all[[pat]] %in% names(friends_genes[[pat]]) |> mean(),
            friends_in_pm = names(friends_genes[[pat]]) %in% pm_all[[pat]] |> mean())
    })
    do.call(rbind, comp)
    ```


Compare with CoGAPS pattern markers again.
    ```{r 'compare with CoGAPS pattern markers again'}
    comp <- lapply(names(pm_all), function(pat){
        list(pattern = pat,
            pm_in_friends = pm_all[[pat]] %in% by_patterns[[pat]] |> mean(),
            friends_in_pm = by_patterns[[pat]] %in% pm_all[[pat]] |> mean()
        )
    })
    do.call(rbind, comp)
    ```

</div>
