---
title: "Find genes that are friendly to a latent pattern (or a cluster) in a matrix"
output: 
  html_vignette:
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
 %\VignetteIndexEntry{Find genes that are friendly to a latent pattern (or a cluster) in a matrix}
---

[Naturam expellas furca, tamen usque recurret]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(rmarkdown)`
`r library(friends.test)`
`r options(width=80,keep.source=TRUE)`
`r options(timeout=300)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`
`r library('fgsea')
`r devtools::load_all()` #remove me
<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">

# Introduction

## Genes that are friends to CoGAPS patterns (or CoGAPS patterns that are friends to genes?)

    ```{r 'setup'}
    data(friends.test.cogaps.example)
    ```

Feature loadings in CoGAPS are normalized expression values across patterns. Each column is a pattern, each row is a gene.

    ```{r 'examine input'}
    featureLoadings <- friends.test.cogaps.example$loadings
    head(featureLoadings)
    ```

Let's find representative genes for all patterns using `friends.test`. We are interested in patterns - friends of genes, so genes need to be in rows, and patterns in columns.

    ```{r 'patterns are genes friends'}
    friends_of_genes <- friends.test(featureLoadings)
    ```

\code{friends.test} returns a list of lists, the first level is the marker (rowname in the table), the second is the friend (colname) and the leave in a vector of the marker index in the initial table, the friend index and the rank of the friend.


    ```{r 'display all genes per friend'}
    friends_df <- friends_of_genes |>
        purrr::flatten() |>
        purrr::reduce(.f=rbind) |> 
        as.data.frame()

    #create a list of friends (gene names) per pattern with purrr
    friends_list <- friends_df |>
        split(friends_df$friend) |> 
        purrr::map(~rownames(featureLoadings)[.$marker]) |>
        setNames(featureLoadings |> colnames())

    ```

Lets examine hallmarks associated with patterns. We can use `fgsea` for that. We will use the loadings as a ranking statistic, and we will look up the hallmark gene sets from MSigDB.

    ```{r 'look up signatures'}
library(fgsea)
library(msigdbr)
msigdbr_df <- msigdbr(species = "human", category = "H")

hallmarks <- lapply(names(friends_list), function(pat) {
    fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol,
                                      msigdbr_df$gs_name),
                    stats = featureLoadings[friends_list[[pat]], pat])
    fgseaRes[padj < 0.05, c('pathway', 'padj')][order(padj)]

}) |> setNames(names(friends_list))

hallmarks

    ```

HERE! 

Compare all pattern friends to CoGAPS pattern markers. All CoGAPS pattern markers are friends, but not all friends are CoGAPS pattern markers.

    ```{r 'compare with CoGAPS pattern markers'}
    pm_all <- friends.test.cogaps.example$markers_all[["PatternMarkers"]]
    comp <- lapply(names(pm_all), function(pat) {
        list(pattern = pat,
            pm_in_friends = pm_all[[pat]] %in% names(friends_genes[[pat]]) |> mean(),
            friends_in_pm = names(friends_genes[[pat]]) %in% pm_all[[pat]] |> mean())
    })
    do.call(rbind, comp)
    ```


Compare with CoGAPS pattern markers again.
    ```{r 'compare with CoGAPS pattern markers again'}
    comp <- lapply(names(pm_all), function(pat){
        list(pattern = pat,
            pm_in_friends = pm_all[[pat]] %in% by_patterns[[pat]] |> mean(),
            friends_in_pm = by_patterns[[pat]] %in% pm_all[[pat]] |> mean()
        )
    })
    do.call(rbind, comp)
    ```

</div>
