---
title: "Find genes that are friendly to a latent pattern (or a cluster) in a matrix"
output: 
  html_vignette:
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
 %\VignetteIndexEntry{Find genes that are friendly to a latent pattern (or a cluster) in a matrix}
---

[Naturam expellas furca, tamen usque recurret]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(rmarkdown)`
`r library(friends.test)`
`r options(width=80,keep.source=TRUE)`
`r options(timeout=300)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`
`r library('fgsea')
`r devtools::load_all()` #remove me
<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">

# Introduction

## Genes that are friends to CoGAPS patterns (or CoGAPS patterns that are friends to genes?)

    ```{r 'setup'}
    data(friends.test.cogaps.example)
    ```

Feature loadings in CoGAPS are normalized expression values across patterns. Each column is a pattern, each row is a gene.

    ```{r 'examine input'}
    featureLoadings <- friends.test.cogaps.example$loadings
    head(featureLoadings)
    ```

Let's find representative genes for all patterns using `friends.test`. We are interested in gene-friends for all patterns, so patterns need to be in rows, and genes in columns.

    ```{r 'friends are genes'}
    friends_genes <- friends.test(t(featureLoadings))
    ```

There are so many friends for each pattern!

    ```{r 'friends summary'}
    lapply(friends_genes, length)
    ```

Now let's inspect loadings values for Pattern4 friends vs non-friends. Higher loading for friend genes is observed, but there is no clear cut-off. 
In the highest loading value decile, about 50% of genes are friends, but also in the lowest decile, there are about 2% of friend genes.

    ```{r 'inspect pattern 4 friends'}
    loadings <- featureLoadings[, "Pattern_4"]
    df_ecdf <- ecdf(loadings)
    df <- data.frame(value = loadings,
        friend = names(loadings) %in% 
            names(friends_genes$Pattern_4),
        loading_ecdf = df_ecdf(loadings)
    )
    aggregate(friend ~ round(loading_ecdf, 1), data = df, 
        FUN = function(x) c(share = mean(x), n = sum(x))
    )
    ```

Lets examine hallmarks associated with Pattern 4 friends. We can use `fgsea` for that. We will use the loadings as a ranking statistic, and we will look up the hallmark gene sets from MSigDB.
```{r 'look up signatures'}
library(fgsea)
library(msigdbr)
msigdbr_df <- msigdbr(species = "human", category = "H")
query <- loadings[]

fgseaRes <- fgsea(pathways = split(msigdbr_df$gene_symbol, msigdbr_df$gs_name),
                   stats = loadings)
fgseaRes[padj < 0.05,c('pathway', 'padj')][order(padj)]
```

Compare all pattern friends to CoGAPS pattern markers. All CoGAPS pattern markers are friends, but not all friends are CoGAPS pattern markers.
```{r 'compare with CoGAPS pattern markers'}
pm_all <- friends.test.cogaps.example$markers_all[["PatternMarkers"]]
comp <- lapply(names(pm_all), function(pat) {
    list(pattern = pat,
        pm_in_friends = pm_all[[pat]] %in% names(friends_genes[[pat]]) |> mean(),
        friends_in_pm = names(friends_genes[[pat]]) %in% pm_all[[pat]] |> mean())
})
do.call(rbind, comp)
```

Lets find very specific genes for each pattern (genes that have only one friend). We can call them "pattern-specific genes".
    ```{r 'fiends are patterns'}
    friends_patterns <- friends.test(featureLoadings, max.friends.n = 1)
    qq <- lapply(friends_patterns,
        function(x) names(x)
    ) |>
    as.data.frame() |>
    t()

    by_patterns <- lapply(unique(qq), function(pat) rownames(qq)[qq == pat])
    names(by_patterns) <- unique(qq)

    by_patterns
    ```

Compare with CoGAPS pattern markers again.
    ```{r 'compare with CoGAPS pattern markers again'}
    comp <- lapply(names(pm_all), function(pat){
        list(pattern = pat,
            pm_in_friends = pm_all[[pat]] %in% by_patterns[[pat]] |> mean(),
            friends_in_pm = by_patterns[[pat]] %in% pm_all[[pat]] |> mean()
        )
    })
    do.call(rbind, comp)
    ```

</div>