---
title: "Find genes that are friendly to a latent pattern (or a cluster) in a matrix"
output: 
  html_vignette:
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
 %\VignetteIndexEntry{Find genes that are friendly to a latent pattern (or a cluster) in a matrix}
---

[Naturam expellas furca, tamen usque recurret]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(rmarkdown)`
`r library(friends.test)`
`r options(width=80,keep.source=TRUE)`
`r options(timeout=300)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`

<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">

# Introduction
 
CoGAPS decomoses an expression dataset into pattetns of expression, ideally, eah of them represent an underlying biological process. We use the [PDAC dataset](https://zenodo.org/records/7709664) from the [Johnson _et al_, 2023](https://www.nature.com/articles/s41596-023-00892-x) as an example. 

    ```{r 'setup'}
    data(friends.test.cogaps.example)
    ```

Feature loadings in CoGAPS are normalized expression values across patterns. Each column is a pattern, each row is a gene. 

    ```{r 'examine input'}
    featureLoadings <- friends.test.cogaps.example$loadings
    head(featureLoadings)
    ```


# Friends test application to the loading matrix

Let's look for marker genes for the patterns using `friends.test`. In this setup, genes (possibly, markers) are rows, patterns (possibly, friends for the markers) are columns.

## freinds.test

We use here the default version of `friends.test` that selects markers by the KS test. As far ad there is ~15000 genes, we use Bonferroni p=value adjustment rather than the default "BH". 

    ```{r 'patterns are genes friends'}
    friends_of_genes <- friends.test(
        featureLoadings,
        p.adjust.method = "bonferroni"
    )
    ```

## Return structure

The return is a list of lists, the outer list is by gene that is a marker, the inner is by pattern that is the marker's friend. The output format allows a marker to have a lot of friends. 
The leave is a 3-integer vector, are the marker index and the friend index in the matrix we started with, and the rank of the friend for the marker. Let's look at the first element, it's rank is 1 as far as it is the first.

    ```{r show list element}
    friends_of_genes[[1]][[1]]
    ```

The names of the list elements are in accordance.
The first marker name:

    ```{r marker name}
    rownames(featureLoadings)[friends_of_genes[[1]][[1]]["marker"]]
    names(friends_of_genes)[1]
    ```

The name of the first friend of the first marker:
    ```{r friend name}
    colnames(featureLoadings)[friends_of_genes[[1]][[1]]["friend"]]
    names(friends_of_genes[[1]])[1]
    ```

The format is flexible enough, we exemplify posiible transformations [later](#the-result-format-transromations).

## Ranks of a marker in its friends and non-friends

Let's look what happens in the friends.test. First, we rank all the genes inside the loadings matrix.

    ```{r ranking of loadings}
    ranks <- row.int.ranks(featureLoadings)
    head(ranks)
    ```

The ranks of the first marker in all he patterns:
    ```{r demo marker 1 all ranks}
    marker_1 <- names(friends_of_genes)[1]
    ranks[marker_1, ]
    ```

The ranks of the first in all the friend patterns:
    ```{r demo marker 1 friend ranks}
    ranks[marker_1, names(friends_of_genes[[1]])]
    ```

The model splitted the patterns into two sets: Pattern_1 where the marker has a very low rank, and the reaminder. 


## Exclusive marker search

 To look for genes that have only one friend (call them exclusive markers), so we use `max.friends.n = 1`. The parameter actually is involved after the main part of the calculation, it filters out the genes that has more than `max.friends.n` friends, these genes are not stored in the output.

    ```{r 'patterns are only best genes friends'}
    best_friends_of_genes <- friends.test(
        featureLoadings,
        p.adjust.method = "bonferroni",
        max.friends.n = 1
    )
    names(best_friends_of_genes)
    ```

In this super=restrictive mode, only one exclusve marker, `r names(best_friends_of_genes)`, is found. Its best frient is the `r names(best_friends_of_genes[[1]])`.

    ```{r demo best friend and ranks}
    the_marker <- names(best_friends_of_genes)[1]
    ranks[the_marker, ]
    ```

That is what the friends.test is a unique test to detect: the `r names(best_friends_of_genes)` has a moderate rank in `r names(best_friends_of_genes[[1]])`, but it is the best of its ranks in all other patterns and it is significantlu higher than all of them.


# Gene set enrichment

To perform an example of gene set enrichment analysis, we use the Baesian version of the test, `friends.test.bic`. We also restrict the number of friends for a marker by 4, a half of number of the patterns, to keep the similar "direction".

    ```{r 'patterns are good genes friends'}
    good_friends_of_genes <- friends.test.bic(
        featureLoadings,
        prior.to.have.friends = 1E-4,
        max.friends.n = 4
    )
    ```

## The result format transromations

For the gene overrrepresentation analysis, we need to gen a list of markers (genes) for each pttern. So, we swap the levels of the list, and we do not need leaves. Here, we can reverse the list via a data frame. It works ok if the number of elements do not reach hundred of thousands. 

    ```{r 'display all genes per best friend'}
    good_friends_df <- good_friends_of_genes |>
        purrr::flatten() |>
        purrr::reduce(.f = rbind) |>
        data.frame() |>
        tibble::remove_rownames() |>
        dplyr::mutate(
            gene = rownames(featureLoadings)[marker],
            pattern = colnames(featureLoadings)[friend]
        )
    head(good_friends_df)
    ```

Now, let's prepare a list of markers by pattern.

    ```{r 'make exclusive marker list'}
    good_markers_list <- good_friends_df |>
        split(good_friends_df$pattern) |>
        purrr::map(~ .x |>
                dplyr::pull(gene) |>
                unique() |>
                sort()
        )
    ```



## Prepare the gene collections

As far as we have gene lists, let's examine them for overreprentation in the gene collections from [MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp). We use `fgsea` and `msigdbr` for that.

First, prepare the collections sets for H (Hallmarks) and for for C4:3CA (Curated Cancer Cell Atlas).

    ```{r 'prepare msigdb collections'}
    msigdbr_H_df <- msigdbr::msigdbr(species = "human", collection = "H")
    collections_H <- split(
        msigdbr_H_df$gene_symbol,
        msigdbr_H_df$gs_name
    )
    collections_H <- split(
        msigdbr_H_df$gene_symbol,
        msigdbr_H_df$gs_name
    )
    msigdbr_C4_3CA_df <-
        msigdbr::msigdbr(
            species = "human",
            collection = "C4",
            subcollection = "3CA"
        )
    collections_C4_3CA <- split(
        msigdbr_C4_3CA_df$gene_symbol,
        msigdbr_C4_3CA_df$gs_name
    )
    ```

Now, let's test the gene enrichment of te exclusive markers for each pattern in the Curated Cancer Cell Atlas and in Hallmarks MSigDB collections.


    ```{r 'C4_3CA signatures in markers'}
    C4_3CA_by_markers <- good_markers_list |>
        purrr::map(\(markers) {
            fgsea::fora(
                pathways = collections_C4_3CA,
                markers,
                rownames(featureLoadings)
            ) |>
                dplyr::filter(padj < 0.05) |>
                dplyr::select(pathway, padj)
        })
    C4_3CA_by_markers
    ```


    ```{r 'Hallmarks signatures in markers'}
    hallmarks_by_markers <- good_markers_list |>
        purrr::map(\(markers) {
            fgsea::fora(
                pathways = collections_H,
                markers,
                rownames(featureLoadings)
            ) |>
                dplyr::filter(padj < 0.05) |>
                dplyr::select(pathway, padj)
        })
    hallmarks_by_markers
    ```


</div>
