---
title: "Find genes that are friendly to a latent pattern (or a cluster) in a matrix"
output: 
  html_vignette:
vignette: >
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
---

[Naturam expellas furca, tamen usque recurret]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(rmarkdown)`
`r library(friends.test)`
`r library(CoGAPS)`
`r options(width=80,keep.source=TRUE)`
`r options(timeout=300)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`
<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">

# Introduction

## Genes that are friends to CoGAPS patterns (or CoGAPS patterns that are friends to genes?)

```{r}
# get CoGAPS result from zenodo
fname <- 'cogapsresult.Rds'
if(!file.exists(fname)){
    download.file('https://zenodo.org/records/7709664/files/cogapsresult.Rds', destfile = fname, mode = 'wb')
}
cg <- readRDS('cogapsresult.Rds')

# get feature loadings
loadings <- cg@featureLoadings
```

Feature loadings in CoGAPS are normalized expression values across patterns. Each column is a pattern, each row is a gene.
```{r}
head(loadings)
```

Let's find representative genes all patterns using `friends.test`. We are interested in all gene-friends for all patterns, so patterns need to be in rows, and genes in columns.
```{r}
friends_genes <- friends.test::friends.test(t(loadings))
```

Wow, there are so many friends for each pattern!
```{r}
aggregate(friend ~ marker, data=friends_genes, length)
```

Inspect loadings vallues for Pattern4 friends vs non-friends. Higher loading for friend genes is observed, but there is no clear cut-off.
```{r}
df <- loadings[,'Pattern_4']
df <- data.frame(value=df, friend=names(df) %in% friends_genes$friend[friends_genes$marker=="Pattern_4"])
plot(df$value, col=df$friend+1, pch=16, xlab='Pattern_4 loading', ylab='Friend status') 
aggregate(value ~ friend, data=df, mean)
aggregate(value ~ friend, data=df, function(x) c(mean=mean(x), sd=sd(x), min=min(x), max=max(x), n=length(x)))
```

Compare all pattern friends to CoGAPS pattern markers. All CoGAPS pattern markers are friends, but not all friends are CoGAPS pattern markers.
```{r}
pm <- patternMarkers(cg)
comp <- lapply(names(pm$PatternMarkers), function(pat){
  list(pattern=pat,
       pm_in_friends=pm$PatternMarkers[[pat]] %in% friends_genes$friend[friends_genes$marker==pat] |> mean(),
       friends_in_pm=friends_genes$friend[friends_genes$marker==pat] %in% pm$PatternMarkers[[pat]] |> mean())
})
do.call(rbind, comp)

```

What happens if we don't transpose?
```{r}
friends_patterns <- friends.test::friends.test(loadings, max.friends.n = 1)

qq <- lapply(friends_patterns, function(x) names(x)) |> as.data.frame() |> t()

by_patterns <- lapply(unique(qq), function(pat) rownames(qq)[qq==pat])

names(by_patterns) <- unique(qq)

by_patterns
```
</div>