---
title: "In a matrix, we look for columns that are friendly to a row"
output: 
  html_vignette:
vignette: >
 %\VignetteIndexEntry{In a matrix, we look for columns that are friendly to a row}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
 %\VignetteDepends{devtools}
---
<!--  tried below html_vignette, did no work
      theme: Nonuniformity test
      toc: true
      toc_depth: 2 -->


[SI AUGUSTUS CERNATUR, CERNANTUR QOUQUE AMICI]{style="float:right"}

`r format(Sys.Date(), "%d %b %Y")` 

`r library(knitr)`
`r library(markdown)`
`r library(friends.test)`
`r options(width=80,keep.source=TRUE)`
`r opts_chunk$set(echo=TRUE,message=FALSE,cache=FALSE,warning=FALSE)`
`r set.seed(42)`
<!-- cache is off, we save all the necessary things in Rda -->

<div style="line-height: 1.5em;">


## Introduction

There is a simple intuition of what it means to be a friend. A friend of Augustus cares about Augustus more than about other people. And, if we see Augustus, then we infer to see friends(s) of Augustus also. Please refer to [our Arxive publication](https://arxiv.org/abs/2512.24843) for more formal introduction.

\paragraph{Introduction} 

Let's picture a set of expression pattern, each represents a biological process. Techically, each pattern is a vector of expression loads of the involved genes.

Sometimes, the expression of a single gene in a sample indicates the activity of the entire pattern there. The simplest case: the gene has a nonzero load in only one pattern.

Actually, the gene may have several nonzero loads, but all of them but the one of interest are relatively small. Then the gene (AKA Augustus) is the marker gene for the pattern, and the marked pattern is the a friend of this gene. 

There could be more then one friend for a marker. In this case, the marker is more important for the friends than for the non-friends.

\paragraph{The test} 

We have two sets:T (rows) and C (columns) and A real matrix A(t,c) that describes the strength of relation between each t and each c; t is an element of T and c is an element of C.

The model applies to many problems in bioinformatics and statistics.

        Example                           $\left\{T\right\}$  $\left\{C\right\}$                    $A\left(t,c\right)$
--------------------------------------    -------------     ----------------------------------  ----------------------------
gene regulation by TFs                    gene              genes under the TF regulation       strength of regulation
fuzzy clustering                          object            cluster                             object weight in cluster
transcription decomposition               transcript        transcription pattern               transcript's load in pattern
weighted graph                            vertex            another vertex                      weight of edge between column and row
--------------------------------------    -------------     ----------------------------------  ----------------------------



\paragraph{Goal} For each row, we want to identify whather there are columns(s) that particularly prefer(s) the row, and list the columns. Then such a column is a friend (or the best friend if there is only one) for the row. We intend to express the meaning of the word "particularly" in this context by a statistical test, that is the **friends.test** provided by this package.

Let's define the relevance of a row to a column as the rank of the $i$-th rows's relation with $j$-the column inside the column, or, in our notation, the rank of $A_{ij}$ in $A_{*j}$, and denote it as $r(t_i,c_j)$. 

The main idea of the **friends.test**  test is that if a row $t_i$ is not preferred by any column(s), the distribution of the ranks of the row in all the columns $r(t_i,c_j)$ behaves like $j$ i.i.d integers that are uniformely distributed in $1..n$. 

## Examples

### Toy example: Bayesian option

    ```{r source, echo=FALSE}
    data.digits <- 2
    p.val.digits <- 4
    ```

Let's illustrate the method with something supersimple.

    ```{r}
    regulation <-
        structure(
            c(
                0.2, 0.25, 0.1, 0.23, 0.3, 0.12, 0.14,
                4, 3, 6, 1, 9, 7, 10,
                3, 1, 5, 6, 3, 4, 11
            ),
            dim = c(7, 3),
            dimnames =
            list(
                c(
                    "Gene1", "Gene2", "Gene3",
                    "Gene4", "Gene5", "Gene6",
                    "Gene7"
                ),
                c("TF1", "TF2", "TF3")
            )
        )
    noquote(regulation)
    ```
What we want to identify is Gene2-TF1 pair. Gene2 is not the strongest downstream gene in the TF1 column (column), it is only the second, but, on the other hand its rank in other TF-based columns is much lower.

First, let's apply the Bayesian version of the method, `best.friends.bic`.

    ```{r}
    friends.test.bic(regulation, prior.to.have.friends = .5, max.friends.n = 1)
    ```

We see the expected Gene2-TF1 pair as well as the Gene1-TF1 pair, that is similar to the expected, Gene1 has rank of 3 in the TF1 column.

The lower is the prior to have friends for a row (it is an obligatory parameter for `friends.test.bic`), the less marker-friend pairs we get.

    ```{r}
    friends.test.bic(regulation, prior.to.have.friends = .33, max.friends.n = 1)

    friends.test.bic(regulation, prior.to.have.friends = .25, max.friends.n = 1)
    ```

In all the three examples above, the `max.friends.n` parameter was set to 1. It means that the function filters out all the rows that have more that one friendly column, or in other words, we look only for the best friends. If we relax the restriction, we get more results, and the `rank` that shows the rank of the friend for the marker, can differ from 1:

    ```{r}
    friends.test.bic(regulation, prior.to.have.friends = .25)
    ```

Note that `max.friends.n = "all"` is the default value for both `friends.test` (see below) and `friends.test.bic`.

### The same toy example: the KS option

Let's run the p-value-based option of the method on the same data. It is less sensitive on small dataset, so we use unpractical threshold for the p-value to see the result.

    ```{r}
    friends.test(regulation, threshold = .5, p.adjust.method = "no",
                 max.friends.n = 1)
    ```

And, as usual, if we relax the `max.friends.no` restriction, we will see more pairs, but now the friends are not required to be unique for a marker.

    ```{r}
    friends.test(regulation, threshold = .5, p.adjust.method = "no")
    ```

### Iris example
Let's look for anomalies in the traditional **iris** dataset: 
    ```{r}
    data("iris")
    print(head(iris, n = 5))
    ```

We use only numeric columns:

    ```{r}
    data("iris")
    friends.test.bic(iris[1:4], prior.to.have.friends = .001)
    ```

So, we see that the flower #14 is important for (we also can think about it as "popular in") the Sepal.Width nomination. Let's understand it a simple request: what is the quantile of the flower #14 in the four columns.

    ```{r}
    apply(iris[, 1:4], 2, function(col) ecdf(col)(col[14]))
    ```

So, flower #14 is in top 50% in Sepal.Width and in lowest 5% for all other sizes. It is exacly what we describe as "Sepal.Width is friend for flower #14 and the flower #14 is a marker for Sepal.Width".

Let's note that the four sizes have different distributions; still, the friends.test naturally works with this type of data.

## Output

\code{friends.test} and \code{friends.test.bic} return a list of lists, the first level is the marker (rowname in the table), the second is the friend (colname) and the leave in a vector of the marker index in the initial table, the friend index and the rank of the friend. Sometimes, you need a data.frame.



    ```{r 'data.frame convesion'}
        friends.test.bic(regulation, prior.to.have.friends = .25) |>
        purrr::flatten() |>
        purrr::reduce(rbind) |>
        as.data.frame() |>
        tibble::remove_rownames()
    ```



## Rakning and random numbers

The result of both `friends.test` and `friends.test.bic` functions depends on random state generator. The dependency is because the test starts with ranking of the elemants of the input matrix in the columns. The test requires the rank values to be integer, so the ties are resolved randomly. It you want a reproduceablle result, keep the initial random generator state. Another option is to rank the matrix bfore the test call and then to pass the rank matrix as the input. It does not consume a lot of additional time, and the renking of already-ranked does not depend on random seeds. The special case when this approach is the only is if some additional information is used in the ranking.

## Parallel calculation

Both main functions, `friends.test` and `friends.test.bic`, carries `purrr::in_parallel` code, so if you want each of them to run in parallel mode, say `mirai::deamons(n)` (n is number of processes) in your R code before the finction call, and `mirai::deamons(0)` after. See [mirai] for more information.

## Session info

    ```{r}
    sessionInfo()
    ```

</div>
